$p$-adic precision

### 2014-02-19 padicprec.tex
Work on loc. anal.

In addition to allowing for more efficient computation of precision, storing the precision separately from the
approximation makes it easier to determine the precision of the result when step-by-step precision tracking
does not reflect the actual precision of the result.

The SOMOS 4 sequence \cite{somos:89a} is defined by the recurrence
\[
u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.
\]
We shall consider the case where the initial
terms $u_0$, $u_1$, $u_2$ and $u_3$ lie in $\Zp^\times$ and have precision $O(p^N)$.

Let us first examine how the absolute precision of $u_n$ varies with $n$
if it is computed from the precision of $u_{n-4}, \ldots, u_{n-1}$ using the recurrence.
The computation of $u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
decreases the precision by a factor $p^{\val(u_n)}$.  Hence the
step-by-step computation returns the value of $u_n$ with precision
\begin{equation}
\label{eq:SOMOS}
O(p^{N-v_n})
\quad \text{with}\quad
v_n = \val(u_0) + \cdots + \val(u_{n-4}).
\end{equation}

On the other hand, one can prove that the SOMOS 4 sequence 
exhibits the \emph{Laurent phenomenon} \cite{fomin-zelevinsky:02a}. This means that, for all integer 
$n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
From the latter formula, it follows directly that if $u_0$, $u_1$,
$u_2$ and $u_3$ are known up to precision $O(p^N)$ then all $u_n$'s
are also known with the same precision. Thus, the term $v_n$ that 
appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
precision but some numerical instability related to the algorithm used to compute $u_n$.

\begin{rmk}
From the above discussion, one can easily derive a numerically stable 
algorithm that computes the SOMOS 4 sequence: 
\begin{enumerate}
\item compute the Laurent polynomials $P_n$ using the induction formula (and working in 
the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$)
\item evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
\end{itemize}
However, computing the $P_n$'s is very time-consuming since it requires 
division in a polynomial ring with $4$ variables and the
size of the coefficients of $P_n$ explodes as $n$ grows.

In Section \ref{ssec:SOMOS-solution}, we shall design an algorithm computing 
the SOMOS 4 sequence which turns out to be, at the same time, efficient 
and numerically stable.
\end{rmk}

@@

\begin{prop} \label{prop : legendre transform}
If $f$ and $g$ are two real convex functions such that $f \geq g,$ then $f^* \leq g^*.$
Furthermore, slopes of $f^*$ corresponds to extreme points of $f$.
\end{prop}

@@

\begin{prop} \label{prop : maj Lambda} \label{prop:locanalytic}
If $m_1$ is a real convex function above $\Lambda(\sum_{n \geq 1}f_n)$.
We define $m_2$ by extending $m_1$ with slope $2$.
Let $\delta$ be the abscissa of the intersection of $m_2$ and the line defined by $y=-\log \varepsilon+ x.$ Then $e^\delta$ is enough to define first order lattices as in lemma  \ref{lem:main}.
\end{prop}
\begin{proof}
$\delta$ is well-defined thanks to corollary \ref{cor : trunc series}, and $m_2 \geq \Lambda(\sum_{n \geq 2}f_n).$

@@

If $x \in B_{\overline{E}}(0,e^\delta),$ $u = \log \vert x \vert,$ then $\Lambda( \sum_{n \geq 2}f_n)(u) \geq -\log \varepsilon+ u.$
With proposition \ref{prop : connexion sup Newton}, for any $m \geq 2,$ $-\log \vert f_m \vert +m u \geq -\log \varepsilon+ u.$

@@

\begin{prop} \label{prop : locanalyticfinitedim}
If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C,$ $\rho$ and $\varepsilon$ are as above. It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $r$ be such that $NP_F^* \left( \log_p r - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $r$ is enough for lemma \ref{lem:main}. 
\end{prop}
\begin{proof}
Same as before, with $L_n=\frac{f^{(n)}(0)}{n!}$ and Legendre formula. The result follows \textit{mutatis mutandis}.
\end{proof}

>>>>

There are also

\todofor{David}{
Explain here why it might be useful to work separately with 
approximation on the one hand and precision on the other hand.
}

\subsection{Step-by-step tracking is not sharp}

In this subsection, we illustrate by some examples that tracking
$p$-adic precision step by step may lead to important numerical
instability.

\subsubsection*{The SOMOS 4 sequence}

The SOMOS 4 sequence is a four-term inductive sequence satisfying:
$$u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.$$
In this example, we shall consider the situation where the initial
terms $u_0$, $u_1$, $u_2$ and $u_3$ are \emph{invertible} $p$-adic 
integers, \emph{i.e.} lie in $\Zp^\times$.

Let us first examine what happens (regarding to the precision) when
we are computing the sequence $(u_n)$ naively. The computation of
$u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
decreases the precision by a factor $p^{\val(u_n)}$ where $\val$
stands for the $p$-adic valuation. Hence, if we assume that $u_0$,
$u_1$, $u_2$ and $u_3$ are given with precision $O(p^N)$ then the
naive computation returns the value of $u_n$ with precision
\begin{equation}
\label{eq:SOMOS}
O(p^{N-v_n})
\quad \text{with}\quad
v_n = \val(u_0) + \cdots + \val(u_{n-4}).
\end{equation}

However, on the other hand, one can prove that the SOMOS 4 sequence 
exhibits the \emph{Laurent phenomenon}. This means that, for all integer 
$n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$ (see \cite{fomin-zelevinsky:02a}).
From the latter formula, it follows directly that if $u_0$, $u_1$,
$u_2$ and $u_3$ are known up to precision $O(p^N)$ then all $u_n$'s
are also known with the same precision. Thus, the term $v_n$ that 
appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
precision but some numerical instability related to the algorithm ---
which includes the model precision --- we used to compute to SOMOS 
sequence.

\begin{rmk}
From the above discussion, one can easily derive a numerically stable 
algorithm that computes the SOMOS 4 sequence: (1)~we compute the 
Laurent polynomials $P_n$ using the induction formula (and working in 
the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$) and (2)~we 
evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
However this algorithm is not efficient at all since computing the 
$P_n$'s is very time-consuming for two reasons: first, it requires 
division in a polynomial ring with $4$ variables and, second, the
size of the coefficients of $P_n$ explodes as $n$ grows.

In \S \ref{ssec:SOMOS-solution}, we shall design an algorithm computing 
the SOMOS 4 sequence which turns out to be, at the same time, efficient 
and numerically stable.
\end{rmk}

@@

\begin{prop} \label{prop : legendre transform}
Slopes of $f^*$ corresponds to extreme points of $f$.
\end{prop}

@@

\begin{prop} \label{prop : maj Lambda} \label{prop:locanalytic}
If $m_1$ is a real convex function above $\Lambda(\sum_{n \geq 1}f_n)$.
We define $m_2$ by extending $m_1$ with slope $2$.
Let $\delta$ be the abscissa of the smallest intersection of $m_2$ and the line defined by $y=-\log \varepsilon+ x,$ or $v_\infty$ if there is no such intersection.  Then $e^\delta$ is enough to define first order lattices as in lemma  \ref{lem:main}.
\end{prop}
\begin{proof}
$\delta$ is well-defined thanks to corollary \ref{cor : trunc series}, and $m_2 \geq \Lambda(\sum_{n \geq 2}f_n).$

@@

By comparing slopes, if $x \in B_{\overline{E}}(0,e^\delta),$ $u = \log \vert x \vert,$  $\Lambda( \sum_{n \geq 2}f_n)(u) \geq -\log \varepsilon+ u.$
With proposition \ref{prop : connexion sup Newton}, for any $m \geq 2,$ $-\log \vert f_m \vert +m u \geq -\log \varepsilon+ u.$

@@

\begin{prop} \label{prop : locanalyticfinitedim}
If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C,$ $\rho$ and $\varepsilon$ are as above. It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $\delta$ be such that $NP_F^* \left( \log_p \delta - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $\delta$ is enough for lemma \ref{lem:main}. 
\end{prop}
\begin{proof}
Same as before, with $L_n=\frac{f^{(n)}(0)}{n!}$ and Legendre formula. The result follows \textit{mutatis mutandis}.
\end{proof}
